On the specific recursive structure of iterated parametricity

### 2025-10-26 paper/paper.tex
paper: minor corrections
A bit further

@@

    \layertype[n][p]                                     & (\framep[n][0,...,p]:\frametype[n][0,...,p]) & : \framep[n][p] \rightarrow \U \\

@@

@@

\begin{align*}
  \deps[n][0,..,p-1] \defeq
  \left(\begin{array}{l}
            \framep[n][0,...,p-1] :\frametype[n][0,...,p-1] \\
            \painting[n][0,...,p-1] :\paintingtype[n][0,...,p-1]
            (\framep[n][0,...,p-1])                         \\
            \restrf[n][0,...,p-1]  :\restrftype[n][0,...,p-1]
            \left(\begin{array}{l}
                \framep[n][0,...,p-1]   \\
                \painting[n][0,...,p-1] \\
              \end{array}\right)                   \\
          \end{array}\right)
\end{align*}
so that $\framep[n][0,...,p]$ has type $\deps[n][0,..,p-1] \rightarrow \frametype[n][0,...,p]$. Later, we shall write $\deps.\framep$, $\deps.\painting$, $\deps.\restrf$ for the corresponding projections of a $\deps$.

@@

To define $\painting$ formally, we need to define $\framep[n][p,...,n]$, $\painting[n][p,...,n]$ and $\restrf[n][p,...,n]$:

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \frametype[n][p,...,n]         &                                              & :         & \U                                                                                       \\
    \frametype[n][]                &                                              & \defeq    & \unittype                                                                                \\
    \frametype[n][p,...,n]         &                                              & \defeq    & \U \times \frametype[n][p+1,...,n-1]                                                     \\
    \\
    \paintingtype[n][p,...,n]      & (\framep[n][p,...,n]:\frametype[n][p,...,n]) & :         & \U                                                                                       \\
    \paintingtype[n][]             & \unitpoint                                   & \defeq    & \unittype                                                                                \\
    \paintingtype[n][p,...,n](p<n) & (\framep[n][p,...,n])                        & \defeq    & (\framep[n][p] \rightarrow \U) \times \paintingtype[n][p+1,...,n](\framep[n][p+1,...,n]) \\
    \\
    \restrftype[n][p,...,n]        &
    \multicolumn{3}{l}{
      \left(\begin{array}{l}
              \deps[n-1][0,...,p-1]:\depstype[n-1][0,...,p-1]                                  \\
              \framep[n-1][p,...,n-1]:\frametype[n-1][p,...,n-1]                               \\
              \painting[n-1][p,...,n-1]:\paintingtype[n-1][p,...,n-1](\framep[n-1][p,...,n-1]) \\
            \end{array}\right) : \U
    }                                                                                                                                                                                    \\
    \restrftype[n][]               &
    \left(\begin{array}{l}
              \deps[n-1][0,...,n-1] \\
              \unitpoint            \\
              \unitpoint            \\
            \end{array}\right)       & \defeq                                       & \unittype                                                                                            \\
    \restrftype[n][p,...,n-1]      &
    \left(\begin{array}{l}
              \deps[n-1][0,...,p-1]     \\
              \framep[n-1][p,...,n-1]   \\
              \painting[n-1][p,...,n-1] \\
            \end{array}\right)
                                   & \defeq                                       &
    \begin{array}{c}
      \framep[n][p]
      \left(\begin{array}{l}
                \deps[n-1][0,...,p-1] \\
              \end{array}\right)
      \rightarrow \framep[n-1][p] \\
      \times                      \\
      \restrftype[n][p+1,...,n]
      \left(\begin{array}{l}
                \deps[n-1][0,...,p]         \\
                \framep[n-1][p+1,...,n-1]   \\
                \painting[n-1][p+1,...,n-1] \\
              \end{array}\right)
    \end{array}
  \end{array}
\end{equation*}

This allows to define $\painting[n][p]$ formally:
\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \painting[n][p]          &
    \multicolumn{3}{l}{
      \left(\begin{array}{l}
                \deps[n-1][0,...,n-1]:\depstype[n-1][0,...,n-1] \\
                E:\framep[n-1][n-1]\rightarrow \U               \\
              \end{array}\right) : \framep[n-1][n-1]\rightarrow \U
    }                                                            \\
    \painting[n][n]          &
    \left(\begin{array}{l}
              \deps[n-1][0,...,n-1] \\
              E                     \\
            \end{array}\right) & \defeq & E                        \\
    \painting[n][p](p<n)     &
    \left(\begin{array}{l}
              \deps[n-1][0,...,n-1] \\
              E                     \\
            \end{array}\right) & \defeq & d \mapsto
    \left(\begin{array}{l}
              \Sigma l:\layer(\deps[n-1][0,...,p-1]). \\
              \painting[n][p+1,...,n-1]
              \left(\begin{array}{l}
                  \deps[n-1][0,...,n-1] \\
                  E                     \\
                \end{array}\right)(d,l)           \\
            \end{array}\right) \\
  \end{array}
\end{equation*}

From which we can deduce $\painting[n][0,...,p]$ by recursion on $p$.

At this stage, we are able to define $\framep$, $\painting$ and $\restrftype$ from $\deps$ at the previous level. To go further, we need to define the sequence $\restrptype$ of types of painting restrictions:

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \restrptype[n][0,...,p] & (\deps[n][0,...,n]:\depstype[n][0,...,n]) & :      & \U                                        \\
    \restrptype[n][]        & \unitpoint                                & \defeq & \unittype                                 \\
    \restrptype[n][0,...,p] & (\deps[n][0,...,n])                       & \defeq & \restrptype[n+1][0,...,p-1] \times \left(
    \begin{array}{l}
        \deps[n+1][0,...,p-1].                            \\ \Pi q\leq n.\Pi \epsilon.\Pi d:\framep(\deps).\\
        \painting[n][p](\deps[n][0,...,n])(d) \rightarrow \\ \deps.\painting[n][p](\deps.\restrf[n][p][q][\epsilon](d))\end{array}
    \right)                                                                                                                  \\
  \end{array}
\end{equation*}

Mutual definition of $\restrf$ and $\cohftype$

@@

@@

                \deps.\frame[n][0,...,p]    \\

@@

    \restrf[n][0]                              &

@@

    \restrf[n][0,...,p+1]                      &

@@

              \restrf[n+1][0,...,p]

@@

              \lambda q.\lambda \epsilon.\lambda (d,l]). \\
              \left(\restrf[n][0,...,p]

@@

                  \end{array}\right)(d), \restrl[n][p]

@@

                    \layer[n+1][p]    \\

@@

      \restrf[n][p][q][\epsilon](\restrf[n+1][p][r][\omega]

@@

      \restrf[n][p][r][\omega](\restrf[n+1][p][q+1][\epsilon]

@@

@@

                                               & :      & \Pi q \leq n. \Pi \epsilon. \Pi d: \framep[n][p].\left(

@@

            \restrf[n][p](\deps[n][0,...,p])   \\

@@

        \layer[n][p](\deps[n][0,...,n])(\restrf[n][p](d))

@@

    \restrl[n][p][q][\epsilon]                 &

@@

@@

@@

    \lambda\omega.\,\overrightarrow{\cohf[n][p][q][q][\epsilon][\omega]}{\restrp[n][p](l \omega)}                          \\

@@

At this new stage, it is convenient to use an abbreviation to combine sequences $\deps[n][0,..,n]$, $\restrp[n][0,...,p-1]$ and $\cohf[n][0,...,p-1]$ in one block, which we call $\depscohs[n][0,...p-1]$:

\begin{align*}
  \depscohstype[n][0,..,p-1] \defeq
  \left(\begin{array}{l}
            \deps[n][0,...,n] :\depstype[n][0,...,n]        \\
            \restrp[n][0,...p-1]: \restrptype[n][0,...,p-1] \\
            \cohf[n][0,...,p-1]: \cohftype[n][0,...,p-1]    \\
          \end{array}\right)
\end{align*}
so that the type of $\restrf[n][0,...,p]$ can be rephrased as $\depscohs[n][0,..,p-1] \rightarrow \restrftype[n][0,...,p]$. Later, we shall write $\depscohs.\deps$, $\depscohs.\restrp$, and $\depscohs.\cohf$ for the corresponding projections of a $\depscohs$.

@@

to do : def of cohptype

Definition of $\cohf[n][p]$, using $\UIP$ to solve the dependency in $\coht[n][p]$.

@@

      (\cohf[n+1][0,...,p], \lambda q.\lambda r.\lambda \epsilon.\lambda \omega.\lambda (\cohf[n][p],\cohl[n][p])). \\
      \left(\cohf[n][0,...,p]

@@

                \end{array}\right), \restrl[n][p]

@@

                  \layer[n+1][p]        \\
                \end{array}\right) (l)\right)                                                                           \\

@@

              \restrp[n][0,...,p-1]:\restrptype[n][0,...,p-1](\depscohs[n][0,...,p-1]) \\

@@

      \cohf[n][p][q][\epsilon](\cohf[n+1][p][r][\omega]

@@

      \cohf[n][p][r][\omega](\cohf[n+1][p][q+1][\epsilon]

@@

@@

    \cohl[n][p][q][\epsilon]                           &

@@

      \Pi d.\Pi l. \overrightarrow{\cohf[n][p][r+1][q+1][\epsilon][\omega]}{\restrl[n][p][q][\epsilon](\restrl[n][p][r][\omega](l))} = \\

>>>>

            \end{array}\right) & :                                            & \U                                                                                                                    \\

@@

    \layertype[n][p]                                                              &
    & :                                     & \Type                        \\
    \layertype[n][p]                                                              &
    & \defeq                                & \framep[n][p]
            \end{array}\right) \rightarrow \U \\

@@

            \end{array}\right)                          & :                                            & \layertype[n][p](\framep[n][0,...,p])                                                            \\

@@

\begin{align*}
  \deps[n][0,..,p-1] \defeq
  \left(\begin{array}{l}
            \framep[n][0,...,p-1] :\frametype[n][0,...,p-1] \\
            \painting[n][0,...,p-1] :\paintingtype[n][0,...,p-1]
            (\framep[n][0,...,p-1])                         \\
            \restrf[n][0,...,p-1]  :\restrftype[n][0,...,p-1]
            \left(\begin{array}{l}
                \framep[n][0,...,p-1]   \\
                \painting[n][0,...,p-1] \\
              \end{array}\right)                   \\
          \end{array}\right)
\end{align*}

so that the type of $\framep[n][0,...,p]$ can be shortened into $\depstype[n][0,..,p-1] \rightarrow \frametype[n][0,...,p]$ and similarly for $\layertype[n][p]$ and $\layer[n][p]$.
Later, we shall write $\deps.\framep$, $\deps.\painting$, $\deps.\restrf$ for the corresponding projections of a $\deps$. By abuse of notation, we also write $\unitpoint$ for the only inhabitant of $\depstype[n][0,..,0-1]$.

@@

We can now formally define $\painting[n][p]$ by recursion on $n-p$:
\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \painting[n][p]          &
    \multicolumn{3}{l}{
      \left(\begin{array}{l}
                \deps[n-1][0,...,n-1]:\depstype[n-1][0,...,n-1] \\
                E:\framep[n-1][n-1]\rightarrow \U               \\
              \end{array}\right) : \framep[n-1][n-1]\rightarrow \U
    }                                                            \\
    \painting[n][n]          &
    \left(\begin{array}{l}
              \deps[n-1][0,...,n-1] \\
              E                     \\
            \end{array}\right) & \defeq & E                        \\
    \painting[n][p](p<n)     &
    \left(\begin{array}{l}
              \deps[n-1][0,...,n-1] \\
              E                     \\
            \end{array}\right) & \defeq & d \mapsto
    \left(\begin{array}{l}
              \Sigma l:\layer(\deps[n-1][0,...,p-1]). \\
              \painting[n][p+1,...,n-1]
              \left(\begin{array}{l}
                  \deps[n-1][0,...,n-1] \\
                  E                     \\
                \end{array}\right)(d,l)           \\
            \end{array}\right) \\
  \end{array}
\end{equation*}

From which we can deduce $\painting[n][0,...,p]$ by a second recursion, this time on $p$:

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \painting[n][0,...,p]          &
    \multicolumn{3}{l}{
      \left(\begin{array}{l}
                \deps[n-1][0,...,n-1]:\depstype[n-1][0,...,n-1] \\
                E:\framep[n-1][n-1]\rightarrow \U               \\
              \end{array}\right) : \paintingtype[n][0,...,p](\framep[n][0,...,p](\deps[n-1][0,...,p]))
    }                                                            \\
    \painting[n][0,...,0-1]          &
    \left(\begin{array}{l}
              \unitpoint \\
              E                     \\
            \end{array}\right) & \defeq & \unittype                        \\
    \painting[n][0,...,p] &
    \left(\begin{array}{l}
              \deps[n-1][0,...,p-1] \\
              E                     \\
            \end{array}\right) & \defeq &
       (\painting[n][0,...,p-1]
       \left(\begin{array}{l}
           \deps[n-1][0,...,n-1] \\
           E                     \\
         \end{array}\right),           \\
       \painting[n][p]
       \left(\begin{array}{l}
           \deps[n-1][0,...,n-1] \\
           E                     \\
         \end{array}\right))           \\
  \end{array}
\end{equation*}

At this stage, we are able to define $\framep$, $\painting$ and $\restrftype$ from $\deps$ at the respective previous level. To go further, we need to define the sequence $\restrptype$ of types of painting restrictions:

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \restrptype[n][0,...,p]                    & (\deps[n][0,...,n]:\depstype[n][0,...,n]) & :      & \U                                        \\
    \restrptype[n][]                           & \unitpoint                                & \defeq & \unittype                                 \\
    \restrptype[n][0,...,p](\deps[n][0,...,n]) & (\deps[n][0,...,n])                       & \defeq & \restrptype[n+1][0,...,p-1] \times \left(
    \begin{array}{l}
        \Pi q\leq n.\Pi \epsilon.\Pi d:\framep(\deps[n][0,...,p]). \\
        \painting[n][p](\deps[n][0,...,n])(d) \rightarrow          \\ \deps.\painting[n][p](\deps.\restrf[n][p][q][\epsilon](d))\end{array}
    \right)                                                                                                                                     \\
  \end{array}
\end{equation*}

Thanks to $\restrptype$, we can mutually define $\restrf$ and $\cohftype$, using abbreviations in passing for $\layer$ and $\layertype$:

@@

                      \restrp[n][0,...,p-2] \\

@@

                \deps.\framep[n][0,...,p]   \\

@@

    \restrf[n][0,...,0-1]                              &

@@

    \restrf[n][0,...,(p+1)-1]                      &

@@

              \restrf[n+1][0,...,p][q][\epsilon]

@@

              \lambda q.\lambda \epsilon.\lambda (d,l). \\
              \left(\restrf[n][p][q][\epsilon]

@@

                  \end{array}\right)(d), \restrl[n][p][q][\epsilon]

@@

@@

      \deps.\restrf[n][p][q][\epsilon](\restrf[n+1][p][r][\omega]

@@

      \deps.\restrf[n][p][r][\omega](\restrf[n+1][p][q+1][\epsilon]

@@

            \end{array}\right) & :      & \U                                                                                   \\
              \restrp[n][0,...,p-1] \\

@@

                                               & \defeq & \Pi q \leq n. \Pi \epsilon. \Pi d: \framep[n][p].\left(

@@

            \restrf[n][p]
              \deps[n][0,...,p]   \\
              \restrp[n][0,...,p-1] \\
              \cohf[n][0,...,p]

@@

        \deps.layer[n][p](\deps[n][0,...,n])(\restrf[n][p]
              \deps[n][0,...,p]   \\
              \restrp[n][0,...,p-1] \\
              \cohf[n][0,...,p]
            \end{array}\right)(d))

@@

    \restrl[n][p]                              &

@@

              \cohf[n][0,...,p-1]:\cohftype[n][0,...,p-1] \\

@@

              \restrp[n][0,...,p-1] \\

@@

    \lambda q\leq n. \lambda \epsilon.\lambda\omega.\,\overrightarrow{\cohf[n][p][q][q][\epsilon][\omega]}{\restrp[n][p](l \omega)} \\

@@

At this new stage, it is convenient to use a new abbreviation to combine sequences of $\deps[n][0,..,n]$, $\restrp[n][0,...,p-1]$ and $\cohf[n][0,...,p-1]$ in one block, which we call $\depscohstype[n][0,...p-1]$:

\begin{align*}
  \depscohstype[n][0,..,p-1] \defeq
  \left(\begin{array}{l}
            \deps[n][0,...,n] :\depstype[n][0,...,n]        \\
            \restrp[n][0,...p-1]: \restrptype[n][0,...,p-1] \\
            \cohf[n][0,...,p-1]: \cohftype[n][0,...,p-1]    \\
          \end{array}\right)
\end{align*}
so that the type of $\restrf[n][0,...,p]$ can be rephrased as $\depscohstype[n][0,..,p-1] \rightarrow \restrftype[n][0,...,p]      \left(\begin{array}{l}
                \deps.\framep[n][0,...,p]   \\
                \deps.\painting[n][0,...,p] \\
              \end{array}\right)$, and similarly for $\restrltype$ and $\restrl$. Later, we shall write $\depscohs.\deps$, $\depscohs.\restrp$, and $\depscohs.\cohf$ for the corresponding projections of an element $\depscohs$ in $\depscohstype$. By abuse of notation, we also write $\unitpoint$ for the only inhabitant of $\depscohstype[n][0,..,0-1]$.

It is also convenient to abbreviate:

\begin{align*}
  \overline{\deps}(\depscohs:\depscohstype) : \depstype\\
  \overline{\deps}(\depscohs) \defeq
\left(\begin{array}{l}
            \framep[n][p](\depscohs.\deps[n][0,...,p])   \\
            \painting[n][p](\depscohs.\deps[n][0,...,n]) \\
            \restrf[n][p](\depscohs[n][0,...,p])\\
      \end{array}\right)\\
\end{align*}

Armed with $\restrframe$ and $\cohftype$, it is now possible to define
$\restrp[n][p]$ by recursion on $n-p$:

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \restrp[n][p]          &
    \multicolumn{3}{l}{
      \left(\begin{array}{l}
                \depscohs[n][0,...,n]:\depscohstype[n][0,...,n] \\
                E:\framep[n-1][n-1]\rightarrow \U               \\
                E':\framep[n][n]\rightarrow \U               \\
      \end{array}\right) & : &
      \Pi q\leq n. \Pi \epsilon. \Pi d:\framep[n][p]
      \left(\begin{array}{l}
                \overline{\deps}(\depscohs)\\
                E':\framep[n][n]\rightarrow \U               \\
      \end{array}\right).\,
      \painting[n][p]
      \left(\begin{array}{l}
                \overline{\deps}(\depscohs)\\
                E':\framep[n][n]\rightarrow \U               \\
      \end{array}\right)(d) \rightarrow
      \painting[n-1][p]
      \left(\begin{array}{l}
                \depscohs[n][0,...,n].\deps\\
                E:\framep[n-1][n-1]\rightarrow \U               \\
      \end{array}\right)(\restrf[n][p](\depscohs[n][0,...,n].deps)(d))
    }                                                            \\
    \restrp[n][n]          &
    \left(\begin{array}{l}
              \depscohs[n][0,...,n] \\
              E                     \\
              E'                    \\
            \end{array}\right) & \defeq & \lambda q\leq n.\lambda \epsilon. \lambda d. \lambda (l,c). l\epsilon                     \\
    \restrp[n][p](p<n)     &
    \left(\begin{array}{l}
              \depscohs[n][0,...,n] \\
              E                     \\
              E'                    \\
            \end{array}\right) & \defeq & \lambda q\leq n.\lambda \epsilon. \lambda d. \lambda (l,c).
    \left(\begin{array}{l}
              \restrl[n][p][q][\epsilon](\depscohs[n-1][0,...,p-1])(l), \\
              \restrp[n][p][q][\epsilon]
              \left(\begin{array}{l}
                  \depscohs[n][0,...,n] \\
                  E                     \\
                  E'                    \\
                \end{array}\right)(\restrf[n][p](\depscohs[n][0,...,p])(d))(c)           \\
            \end{array}\right) \\
  \end{array}
\end{equation*}

From which we can deduce $\restrp[n][0,...,p]$ again by a second recursion, this time on $p$:

...

@@

At this stage, we are then able to define $\restrf$, $\restrp$ and $\cohftype$
from $\depscohs$ at the previous level. To go further, we need to define the
sequence $\cohptype$ of types of painting coherence laws:

\begin{equation*}
  \fontsize{7.2}{9}\selectfont
  \begin{array}{llcl}
    \cohptype[n][0,...,p]                    & (\depscohs[n][0,...,n]:\depscohstype[n][0,...,n]) & :      & \U                                        \\
    \cohptype[n][]                           & \unitpoint                                & \defeq & \unittype                                 \\
    \cohptype[n][0,...,p](\depscohs[n][0,...,n]) & (\depscohs[n][0,...,n])                       & \defeq & \cohptype[n+1][0,...,p-1] \times \left(
    \begin{array}{l}
        \Pi q\leq n.\Pi \epsilon.\Pi d:\framep(\depscohs[n][0,...,p]). \\
        \painting[n][p](\depscohs[n][0,...,n])(d) \rightarrow          \\ \depscohs.\painting[n][p](\depscohs.\restrf[n][p][q][\epsilon](d))\end{array}
    \right)                                                                                                                                     \\
  \end{array}
\end{equation*}

This is enough to define $\cohf[n][p]$, using $\UIP$ to solve the dependency in $\coht[n][p]$.

@@

      (\cohf[n+1][0,...,p], \lambda q.\lambda r.\lambda \epsilon.\lambda \omega.\lambda (d,l). \\
      \left(\cohf[n][0,...,p][q][r][\epsilon][\omega]

@@

                \end{array}\right)(d), \cohl[n][p][q][r][\epsilon][\omega]

@@

                \end{array}\right) (d)(l)\right)                                                                           \\

@@

              \cohp[n][0,...,p-1]:\cohptype[n][0,...,p-1](\depscohs[n][0,...,p-1]) \\

@@

      \Pi q\leq n.\,\Pi r\leq q.\,\Pi \epsilon.\,\Pi \omega.\,\Pi \theta \\
      \depscohs.\cohf[n][p][r][p][\omega][\theta] \circ \cohf[n+1][p][q+1][p][\epsilon][\theta]

@@

                \coht[n+1][0,...,p-1]
      \circ \depscohs.\cohf[n][p][q][r][\epsilon][\omega] =                         \\
      \cohf[n+1][p][q+1][r+1][\epsilon][\omega]

@@

                \coht[n+1][0,...,p-1]
      \circ \depscohs.\cohf[n][p][q][p][\epsilon][\theta]
      \circ \cohf[n+1][p][r][p][\omega][\theta]
                \coht[n+1][0,...,p-1]
      \end{array}\right)\\

@@

    \cohl[n][p]                           &

@@

      \Pi d.\Pi l. \overrightarrow{\cohf[n][p][r+1][q+1][\epsilon][\omega]}{(\restrl[n][p][q][\epsilon](\restrl[n][p][r][\omega](l)))} = \\
